### 2021-02-24 manuscript/manuscript-speed-accuracy.tex
Major update: larger dataset, revised and simplified code, new figures.

@@

\Authors{Paul P. Gardner\textsuperscript{1,2,3}*, James M. Paterson\textsuperscript{1,2}, Fatemeh Ashari-Ghomi\textsuperscript{1,2}, Sinan U. Umu\textsuperscript{1,2}, Stephanie McGimpsey\textsuperscript{4,5}, Aleksandra Pawlik\textsuperscript{6}} % Authors
\affiliation{\textsuperscript{1}\textit{School of Biological Sciences, University of Canterbury, Christchurch, New Zealand.}} % Author affiliation
\affiliation{\textsuperscript{2}\textit{Biomolecular Interaction Centre, University of Canterbury, Christchurch, New Zealand.}}
\affiliation{\textsuperscript{3}\textit{Bio-Protection Research Centre, University of Canterbury, Christchurch, New Zealand.}}
\affiliation{\textsuperscript{4}\textit{McConomy School of Dance, Derry, Ireland.}}
\affiliation{\textsuperscript{5}\textit{Research for Good, Limavady, Ireland.}}
\affiliation{\textsuperscript{6}\textit{New Zealand eScience Infrastructure, 49 Symonds St, Auckland, New Zealand.}}
\affiliation{*\textbf{Corresponding author}: paul.gardner@canterbury.ac.nz} % Corresponding author

@@

%\Abstract{In the below we provide additional results for our investigation of computational biology benchmarks.}
\Abstract{
Computational biology has provided widely used and powerful software tools for testing and making inferences about biological data. In the face of increasing volumes of data, heuristic methods that trade software speed for mathematical completeness must be employed. 
We are interested in whether trade-offs between speed and accuracy are reasonable. Also, what factors are indicative of accurate software?

In this work we mine published benchmarks of computational biology software, we collect data on the relative accuracy and speed of different software and then test to see what factors influence accuracy e.g. speed, author reputation, journal impact or recency.

We found that author reputation, journal impact, the number of citations, software speed and age are not reliable predictors of software accuracy. Unfortunately, citations, author and journal reputation are frequently cited reasons for selecting software tools. Our results imply that useful bioinformatics software is not only the domain of famous senior researchers. In addition, we found that there is an excess of slow and inaccurate software tools across multiple sub-disciplines of bioinformatics. Meanwhile, there are very few tools of middling accuracy and speed. We hypothesise that a form of publication-bias unduly influences the publication and development of bioinformatic software tools. In other words, at present software that is not highly ranked on speed and not highly ranked on accuracy is difficult to publish due to editorial and reviewer practices. This leaves an unfortunate gap in the literature upon which future software refinements cannot be constructed.
}

@@

\section*{Background}
Computational biology software is widely used and has produced some of
the most cited publications in the scientific corpus
\cite{Perez-Iratxeta2007-lv,Van_Noorden2014-kc,Wren2016-xy}. This
software includes implementations of methods for sequence alignment
and homology inference
\cite{Altschul1990-ht,Thompson1994-eu,Thompson1997-rl,Altschul1997-ga},
phylogenetic analysis
\cite{Felsenstein1985-lj,Saitou1987-zl,Posada1998-qq,Ronquist2003-yh,Tamura2007-ei},
statistical analysis of survival patterns in biomedicine
\cite{Kaplan1958-ju,Cox1972-nu}, biomolecular structure analysis
\cite{Sheldrick1990-kc,Sheldrick2008-xy,Jones1991-ik,Laskowski1993-vi,Otwinowski1997-xj},
visualization and data collection
\cite{Kraulis1991-lt,Berman2000-to}. However, the popularity of a
software tool does not necessarily imply that it is either accurate or
computationally efficient, instead usability, ease of installation,
operating system and other factors may play a greater role. There have
been several notable incidences where inaccurate software has caused
considerable harm \cite{leveson1993investigation}.
%Fergusson covid predictions?

@@

The gold-standard for determining accuracy is for independent
researchers to conduct benchmarks, which can serve a useful role in
reducing the over-optimistic reporting of software accuracy
\cite{Boulesteix2010-te,Jelizarow2010-zf} and the self-assessment trap
\cite{Norel2011-cq}. Benchmark studies typically use a number of
positive and negative control datasets, predictions can then be
partitioned into true or false groups and a variety of metrics can be
used to evaluate the performance of different predictions
\cite{Egan1975-nd,Hall2012-kg}. Some benchmarks now use live, or
frequently updated, results to indicate the latest developments in
software performance
\cite{Bujnicki2001-xr,Puton2014-hy,Barton_undated-er}. The aim of
these benchmarks is to independently identify tools that make
acceptable compromises in terms of scoring schemes and the resulting
potential for false predictions, true predictions and speed, and are
therefore suited for wide adoption by the community.

For common computational biology tasks, a proliferation of
software-based solutions often exists
\cite{Felsenstein1995-ic,Altschul2013-bv,Henry2014-ut,Wikipedia_contributors2015-vj,Wikipedia_contributors2015-hr}. While
this may generally be a good problem to have, and points to a
diversity of options from which practical solutions can be selected,
many possible options creates a dilemma for users. In the absence of
any recent gold-standard benchmarks, how should scientific software be
selected? In the following work, we presume that ``biological
accuracy'' is the most desirable feature of software. Biological
accuracy is the degree to which predictions or measurements reflect
the truths of biological systems, this is usually determined by
comparing results of software to the results of established reference
data. These are commonly expert-derived curated datasets. In some
fields biological accuracy is very difficult to ascertain, for
example, in phylogenetics it is nearly impossible to know the
ancestral relationships between organisms. In situations like this,
researchers can use a mix of simulated or high-confidence datasets.

A number of possible predictors of software quality are used by the
community of computational biology software users. Some accessible,
quantifiable and frequently used proxies for identifying high quality
software include: \textbf{1. Recency:} recently published software
tools may have built upon the results of past work or be an update to
an existing software. Therefore, these could be more accurate and
faster. \textbf{2. Wide adoption:} a software tool may be widely used
because it is fast and accurate, or because it is well-supported and
user-friendly. In fact, “large user base”, “word-of-mouth”,
“wide-adoption”, "personal recommendation," and "recommendation from a
close colleague," are frequent responses to surveys of “how do
scientists select software?”
\cite{Hannay2009-cf,Joppa2013-vj,Loman2015-bw}. \textbf{3. Journal
  impact:} high profile journals are run by editors and reviewers who
carefully select and curate the best manuscripts. Therefore, high
impact journals may be more likely to select manuscripts describing
good software \cite{Garfield1955-wf}. \textbf{4. Author/Group
  reputation:} the key to any project is the skills of the people
involved, including maintaining a high collective intelligence
\cite{Joppa2013-vj,Woolley2010-ld,Cheruvelil2014-xn}. As a
consequence, an argument could be made that well respected and
high-profile authors will produce better software
\cite{Hirsch2005-mt,Bornmann2008-il}. \textbf{5. Speed:} software is
frequently said to trade accuracy for speed. For example, heuristic
software such as the popular homology search tool, BLAST, compromise
the mathematical guarantee of optimal solutions for more speed
\cite{Altschul1990-ht,Altschul1997-ga}. Some researchers may naively
interpret this fact as slower software is likely to be more
accurate. But speed may also be influenced by the programming language
\cite{Fourment2008-vl}, and/or level of hardware optimisation
\cite{Farrar2007-ky,Dematte2010-ph}; In general the implementation is
likely to have more of an impact (e.g. brute-force approaches versus
rapid and sensitive pre-filtering
\cite{Schaeffer1989-mu,Papadimitriou_undated-bo}).

Other factors that are less quantifiable that influence whether a
software tool is selected include: whether the documentation is good,
user-friendly, word-of-mouth and ``used in a similar analysis''
\cite{Loman2015-bw}. This sort of information is not as readily
quantifiable as the above measures. However, citation metrics may be a
useful proxy. The word-of-mouth factor may also explain the reason why
some software continues to be used, in spite of poor relative
performance \cite{Wadi2016-dj}. We note that none of these factors
are directly linked to the top ranked features of successful software
projects, as determined by literature survey \cite{nasir2011critical}.

{\bf EDITS:\\
  Number of releases. GitHub statistics (https://www.frontiersin.org/articles/10.3389/fbioe.2018.00198/abstract).\\
  The influence of organizational structure on software quality https://ieeexplore.ieee.org/iel5/4814109/4814110/04814163.pdf
  }

In the following study, we explore factors that may be indicative of software accuracy. This, in our opinion, should be one of the prime reasons for selecting a software tool. We have mined the large and freely accessible PubMed database \cite{Sayers2010-vm} for benchmarks of computational biology software, and manually extracted accuracy and speed rankings for \textbf{{\color{black}243}} software packages. For each software tool, we have collected measures that may be predictive of accuracy, and may be subjectively employed by the researcher community as a proxy for software quality. These include relative speed, relative age, the productivity and impact of the corresponding authors, journal impact and the number of citations.

\section*{Results}
We have collected relative accuracy and speed ranks for \textbf{{\color{black}243}} distinct software tools. This software has been developed for solving a broad cross-section computational biology tasks. These include software for homology search \cite{Freyhult2007-et}, genome sequence analysis (e.g. read mapping or sequence assembly) \cite{Junemann2014-mb,Tran2014-pe,Zhang2011-nd,Abbas2014-gu,Bao2011-lv,Caboche2014-lj,Kleftogiannis2013-wi,Hatem2013-cs,Schbath2012-ob,Ruffalo2011-rl,Yang2013-aj,Holtgrewe2011-fd,Rackham2015-ag,Huang2015-wu}, multiple sequence alignment \cite{Thompson2011-rf,Nuin2006-nk,Pais2014-sr,Pervez2014-zp,Liu2010-rp}, cell tracking \cite{Maska2014-ak}, transcriptome analysis \cite{Li2012-wr,Lu2013-fs,Liu2014-kz,Kumar2016-xz}, RNA interaction prediction \cite{Pain2015-gr}, protein interactions \cite{Tikk2010-qd}, protein structure prediction \cite{Kolodny2005-ry,Wallner2005-qi}, epistasis \cite{Shang2011-vy},  metagenomic analysis \cite{Lindgreen2016-tt,Bazinet2012-wf}, repetitive sequence prediction \cite{Saha2008-kd}, proteomics \cite{Lange2008-pt,Yang2009-oc} and phylogenetics \cite{Liu2011-pz,Yang2011-dv,Oscamou2008-md,Bayzid2013-hc,Liu2009-lx}. Each software tool was benchmarked in at least one of \textbf{{\color{black}43}} publications that satisfy the Boulesteix criteria \cite{Boulesteix2013-vb}. In brief, the Boulesteix criteria are: 1. the main focus of the article is a benchmark. 2. the authors are reasonably neutral. 3. the test data and evaluation criteria are sensible.

For each of the publications describing these methods, we have (when possible) identified the 2014 journal impact factor (JIF), published by Thomson Reuters \cite{Garfield2006-lv} and the H5-index published by \href{https://scholar.google.co.nz/citations?view_op=top_venues\&hl=en}{Google Scholar Metrics}. We have collected the H-indices and M-indices \cite{Hirsch2005-mt} for the corresponding authors for each method, and the number of times the publication(s) associated with a method has been cited using Google Scholar (data collected over a 1 month period in early 2016).

We have computed the Spearman’s correlation coefficient for each pairwise combination of the mean normalised accuracy and speed ranks, the year published, mean relative age (compared to software in the same benchmarks), journal IF and H5 metrics, the total number of citations, the relative number of citations (compared to software in the same benchmarks) and the maximum H and M indices for the corresponding authors. The results are presented in Figure~\ref{fig:allfactors}A. We found significant associations between most of the citation-based metrics (journal H5, JIF, citations, relative citations, H-index and M-index). There is also a strong association between the year of publication, the relative age and many of the citation-based metrics.

\begin{figure*}[h]
\includegraphics[width=\textwidth]{figure1.pdf}
\caption{\textbf{A.} A heatmap indicating the relationships between proposed predictors of software quality. Spearman’s rho is used to infer correlations between metrics such as the H and M indices of corresponding authors, number of citations, journal impact factors and H5 indices, the year and relative age of software and the mean relative rankings of software speed and accuracy. Red colours indicate a high positive correlation, blue colours indicate a high negative correlation. Correlations with a P-value less than 0.05 are indicated with a ‘X’. The dashed rectangular area is illustrated in more detail in \textbf{B}, the bold square is shown in more detail in \textbf{Figure 2}.The dendrogram was computed using the default ‘heatmap.2 function in R, which computes a Euclidean distance matrix and a complete-linkage hierarchical clustering. \textbf{B.} A barplot illustrating the correlation as measured by Spearman’s rho between and normalised accuracy ranks and potential factors that may be predictive of accuracy. In order to give an appreciation of the difference between the observed effect-sizes and significant effect sizes, we generated 10,000 permuted accuracy ranks for each benchmark and recorded Spearman’s rho for the significant correlations (P<0.05). These values are marked with “x”s in the barplot.}
\label{fig:allfactors}
\end{figure*}

We found that author reputation metrics, journal impacts and the age of methods were \textbf{not} significantly correlated with either method accuracy or speed (see Figure~\ref{fig:allfactors}). The strongest association was between accuracy and journal impact factor (Spearman’s rho = 0.1, P-value = 0.16). A linear model of these parameters and accuracy also failed to identify a correlation between these (accuracy: R2 = -0.03, P-value = 0.94; speed: R2 = 0.04, P-value = 0.66). To further validate this result, we compute a correlation between speed and accuracy for each benchmark and used weighted sum Z-tests \cite{Zaykin2011-tj}. This also failed to identify a significant relationship (sum Z=-0.1, P-value=0.5).

In order to gain a deeper understanding of the distribution of available bioinformatic software tools on a speed versus accuracy landscape, we ran a Monte Carlo permutation test. The ranks extracted from each benchmark were randomly permuted, generating 10,000 randomized speed and accuracy ranks. In the cells of a $10\times10$ grid spanning the normalised speed and accuracy ranks we computed a Z-score for the observed number of methods in a cell, compared to the expected distributions generated by the randomized ranks. The results of which are shown in Figure~\ref{fig:speedaccuracy}. We identified 19 bins where there was a significant excess or dearth of methods. For example, there was an excess of “slow and inaccurate” software (Z=1.6, P-value=0.05) and “slow and accurate” software (Z=1.9, P-value=0.03). We find that the amount of software classed as “fast and accurate” and “fast and inaccurate” are at approximately the expected proportions based upon the permutation test. The number of significant results is in excess and not due to multiple testing, as the probability of finding 19 of 100 tests significant by chance is low (P-value = $5\times 10^{-7}$, exact binomial test).

@@

There is a large reduction in the number of software tools that are classed as intermediate in terms of both speed and accuracy based upon our permutation test (Figure~\ref{fig:speedaccuracy}). The cells corresponding to the four central underrepresented deciles are highlighted (Z = -1.7, -2.1, -2.1 and -2.4, P-values = 0.04, 0.02, 0.02 and 0.008, respectively, reading from top to bottom, left to right). We also tested the relative age of the software tools in significantly over- or under-represented regions of the speed vs accuracy plot. We found that the “slow and inaccurate” methods were generally published earlier than other methods (W = 31, P=0.007, one-tailed Wilcoxon test) (Figure S8).

\section*{Discussion}

We have gathered data on the relative speed and accuracies of \textbf{{\color{black}243}} bioinformatic methods from \textbf{{\color{black}43}} benchmarks that were published between 2005 and 2016. We show that there is an under-representation of software that has both intermediate levels of accuracy and speed. There may be a number of factors that drive this phenomena. One possible explanation for this is that benchmarkers select software that fills the extremes of each accuracy/speed niche. This seems unlikely since based upon our word frequency analysis the terms ``broad'', ``comprehensive'', ``inclusive'' and ``thorough'' are over-represented by between 200\% and 4,400\%.  A more likely explanation, in our opinion, is that bioinformatic software tools suffer from a form of publication-bias \cite{Boulesteix2015-am,Nissen:2016}. That is, the probability of a study being published is influenced by the results it contains \cite{sterling1959publication}. The community of developers, reviewers and editors appears to be unwilling to publish software that is neither the fastest nor the most accurate (we have anecdotal evidence to support this, N=1). If correct, this is unfortunate. Some problems that lack fast and accurate solutions force researchers to make unnecessary compromises in terms of accuracy and time. If our hypothesis that this underrepresentation is due to publication-bias is valid then why is there an enrichment of slow and inaccurate software (P-value=0.05, empirical distributions from permutation tests)? How could these methods be published? We have found that slow and inaccurate software is generally published earlier than alternative methods (P=0.007, one-tailed Wilcoxon test), therefore comparisons were not required to publish these tools and the accuracies may have been over-optimistically reported \cite{Boulesteix2015-nc}.

% 0.91	0.0004054328	0.0002104564	broad             100*(0.0004054328/0.0002104564) = 192.6446
% 2.08	0.0028380296	0.0006656728	comprehensive     100*(0.0028380296/0.0006656728) = 426.34
% 2.23	0.0002027164	0.0000352306	thorough          100*(0.0002027164/0.0000352306) = 575.3987
% 3.86	0.0002027164	0.0000046356	inclusive         100*(0.0002027164/0.0000046356) = 4373.035

>>>>

\def\numTools{499}
%cat meanRankSpeedData.tsv | cut -f 4 | grep -v ^method$ | wc -l
\def\numBenchmarkPubs{69}
%cat speed-vs-accuracy-toolRanks2005-2020.tsv | cut -f 1 | sort  | uniq | grep -E '[0-9]'  | wc -l
\def\numBenchmarks{134}
%cat rawRankSpeedData2005-2020.tsv | cut -f 1 | grep -v ^testId$ | uniq -c | wc -l

@@

\Authors{Paul P. Gardner\textsuperscript{1,2}*, James M. Paterson\textsuperscript{?}, Fatemeh Ashari-Ghomi\textsuperscript{?}, Sinan U. Umu\textsuperscript{?}, Stephanie McGimpsey\textsuperscript{3,4}, Aleksandra Pawlik\textsuperscript{5},
Alex Gavryushkin\textsuperscript{6}, Michael A Black\textsuperscript{1}
} % Authors

\affiliation{\textsuperscript{1}\textit{Department of Biochemistry, University of Otago, Dunedin, New Zealand.}} % Author affiliation
\affiliation{\textsuperscript{2}\textit{Biomolecular Interaction Centre, University of Canterbury, Christchurch, New Zealand.}}
\affiliation{\textsuperscript{3}\textit{McConomy School of Dance, Derry, Ireland.}}
\affiliation{\textsuperscript{4}\textit{Research for Good, Limavady, Ireland.}}
\affiliation{\textsuperscript{5}\textit{New Zealand eScience Infrastructure, 49 Symonds St, Auckland, New Zealand.}}
\affiliation{\textsuperscript{6}\textit{Department of Computer Science, University of Otago, Dunedin, New Zealand.}} % Author affiliation
\affiliation{*\textbf{Corresponding author}: paul.gardner@otago.ac.nz} % Corresponding author

@@

%\Abstract{In the below we provide additional results for our investigation of computational biology benchmarks.}
\Abstract{ Computational biology has provided widely used and powerful
  software tools for testing and making inferences about biological
  data. In the face of rapidly increasing volumes of data, heuristic
  methods that trade software speed for mathematical completeness must
  be employed.  We are interested in testing for trade-offs between
  speed and accuracy and for other factors are indicative of accurate
  software.

We have extracted accuracy and speed ranks from independent benchmarks of
computational biology software of different software tools, and evaluate the
factors that are likely to influence accuracy e.g. speed, author
reputation, journal impact, recency or developer efforts.

We found that software speed, author reputation, journal impact, the
number of citations and age are all unreliable predictors of software
accuracy. This is unfortunate because citations, author and journal
reputation are frequently used reasons for selecting software
tools. Our results do show that accurate bioinformatic software tools
are generally the product many gradual improvements, often from
multiple developers, based upon github records. In addition, we find
that bioinformatics has an excess of slow and inaccurate software
tools across many sub-disciplines. Meanwhile, there are few tools that
are middle-of-road in terms of accuracy and speed trade-offs. We
hypothesise that a form of publication-bias unduly influences the
publication and development of bioinformatic software tools. In other
words, at present software that is not highly ranked on speed and not
highly ranked on accuracy is difficult to publish due to author,
editor and reviewer practices. This leaves an unfortunate gap in the
literature upon which future software refinements cannot be
constructed.  }

@@

\section*{Background}
Computational biology software is widely used and has produced some of
the most cited publications in the entire scientific corpus
\cite{Perez-Iratxeta2007-lv,Van_Noorden2014-kc,Wren2016-xy}. This
highly-cited software includes implementations of methods for sequence
alignment and homology inference
\cite{Altschul1990-ht,Thompson1994-eu,Thompson1997-rl,Altschul1997-ga},
phylogenetic analysis
\cite{Felsenstein1985-lj,Saitou1987-zl,Posada1998-qq,Ronquist2003-yh,Tamura2007-ei},
statistical analysis of survival patterns in biomedicine
\cite{Kaplan1958-ju,Cox1972-nu}, biomolecular structure analysis
\cite{Sheldrick1990-kc,Sheldrick2008-xy,Jones1991-ik,Laskowski1993-vi,Otwinowski1997-xj},
visualization and data collection
\cite{Kraulis1991-lt,Berman2000-to}. However, the popularity of a
software tool does not necessarily mean that it is accurate or
computationally efficient, instead usability, ease of installation,
operating system and other factors may play a greater role in a software
tool's popularity. There have been several notable incidences where
inaccurate software has caused considerable harm
\cite{leveson1993investigation}.
%Fergusson covid predictions?

@@

The gold-standard for determining accuracy is for independent
researchers to conduct benchmarks, which can serve a useful role in
reducing the over-optimistic reporting of software accuracy
\cite{Boulesteix2010-te,Jelizarow2010-zf,Weber:2019} and the self-assessment trap
\cite{Norel2011-cq,Buchka:2021}. Benchmark studies typically use a number of
positive and negative control datasets, predictions from different software tools can then be
partitioned into true or false groups and a variety of metrics can be
used to evaluate the performance differences
\cite{Egan1975-nd,Hall2012-kg,Weber:2019}.
%Some benchmarks now use live, or
%frequently updated, results to indicate the latest developments in
%software performance
%\cite{Bujnicki2001-xr,Puton2014-hy,Barton_undated-er}.
The aim of these benchmarks is to independently identify tools that
make acceptable compromises in terms of balancing speed with
discriminating true and false predictions, and are therefore suited
for wide adoption by the community.

For common computational biology tasks, a proliferation of
software-based solutions often exists
\cite{Felsenstein1995-ic,Altschul2013-bv,Henry2014-ut,Wikipedia_contributors2015-vj,Wikipedia_contributors2015-hr}. While
this is a good problem to have, and points to a
diversity of options from which practical solutions can be selected,
many possible options creates a dilemma for users. In the absence of
any recent gold-standard benchmarks, how should scientific software be
selected? In the following we presume that ``biological
accuracy'' is the most desirable feature for a software tool. Biological
accuracy is the degree to which predictions or measurements reflect
the biological truths based on expert-derived curated datasets. In some
fields biological accuracy is very difficult to ascertain, for
example, in phylogenetics it is nearly impossible to know the
true ancestral relationships between organisms. In situations like this,
researchers may use simulated or high-confidence datasets.

A number of possible predictors of software quality are used by the
community of computational biology software users. Some accessible,
quantifiable and frequently used proxies for identifying high quality
software include: \textbf{1. Recency:} recently published software
tools may have built upon the results of past work or be an update to
an existing software. Therefore, these could be more accurate and
faster. \textbf{2. Wide adoption:} a software tool may be widely used
because it is fast and accurate, or because it is well-supported and
user-friendly. In fact, “large user base”, “word-of-mouth”,
“wide-adoption”, "personal recommendation," and "recommendation from a
close colleague," are frequent responses to surveys of “how do
scientists select software?”
\cite{Hannay2009-cf,Joppa2013-vj,Loman2015-bw}. \textbf{3. Journal
  impact:} high profile journals are run by editors and reviewers who
carefully select and curate the best manuscripts. Therefore, high
impact journals may be more likely to select manuscripts describing
good software \cite{Garfield1955-wf}. \textbf{4. Author/Group
  reputation:} the key to any project is the skills of the people
involved, including maintaining a high collective intelligence
\cite{Joppa2013-vj,Woolley2010-ld,Cheruvelil2014-xn}. As a
consequence, an argument could be made that well respected and
high-profile authors will produce better software
\cite{Hirsch2005-mt,Bornmann2008-il}. \textbf{5. Speed:} software tools
frequently trade accuracy for speed. For example, heuristic
software such as the popular homology search tool, BLAST, compromise
the mathematical guarantee of optimal solutions for more speed
\cite{Altschul1990-ht,Altschul1997-ga}. Some researchers may naively
interpret this fact as slower software is likely to be more
accurate. But speed may also be influenced by the programming language
\cite{Fourment2008-vl}, and the level of hardware optimisation
\cite{Farrar2007-ky,Dematte2010-ph}; however, the implementation is
likely to have more of an impact (e.g. brute-force approaches versus
rapid and sensitive pre-filtering
\cite{Schaeffer1989-mu,Papadimitriou_undated-bo}).

Other factors that are less quantifiable that influence whether a
software tool is selected include: whether the documentation is good,
user-friendly, word-of-mouth and ``used in a similar analysis''
\cite{Loman2015-bw}.
%This information is not as readily
%quantifiable as the above measures. However,
However, citation metrics may be a
useful proxy for the above.
%This may also explain the reason why
%some software continues to be used, in spite of poor relative
%performance \cite{Wadi2016-dj}.
%We note that none of these factors
%are directly linked to the top ranked features of successful software
%projects, as based upon literature surveys \cite{nasir2011critical}.

With the wide adoption of Github ($\approx 47\%$ of the
\textbf{{\color{red}\numTools}} tools included in this study could be
linked to a Github repository), quantifiable data on software
development such as the number of contributors to code, number of code
changes and versions is also available
\cite{ray2014large,Dozmorov:2018,mangul2018comprehensive}.

In the following study, we explore factors that may be indicative of
software accuracy. This, in our opinion, should be one of the prime
reasons for selecting a software tool. We have mined the large and
freely accessible PubMed database \cite{Sayers2010-vm} for benchmarks
of computational biology software, and manually extracted accuracy and
speed rankings for \textbf{{\color{red}\numTools~}} unique software tools. For
each tool, we have collected measures that may be predictive
of accuracy, and may be subjectively employed by the researcher
community as a proxy for software quality. These include relative
speed, relative age, the productivity and impact of the corresponding
authors, journal impact and the number of citations.

\section*{Results}
We have collected relative accuracy and speed ranks for
\textbf{{\color{red}\numTools~}} distinct software tools. This software has
been developed for solving a broad cross-section computational biology
tasks.
%% These include software for homology search
%% \cite{Freyhult2007-et}, genome sequence analysis (e.g. read mapping or
%% sequence assembly)
%% \cite{Junemann2014-mb,Tran2014-pe,Zhang2011-nd,Abbas2014-gu,Bao2011-lv,Caboche2014-lj,Kleftogiannis2013-wi,Hatem2013-cs,Schbath2012-ob,Ruffalo2011-rl,Yang2013-aj,Holtgrewe2011-fd,Rackham2015-ag,Huang2015-wu},
%% multiple sequence alignment
%% \cite{Thompson2011-rf,Nuin2006-nk,Pais2014-sr,Pervez2014-zp,Liu2010-rp},
%% cell tracking \cite{Maska2014-ak}, transcriptome analysis
%% \cite{Li2012-wr,Lu2013-fs,Liu2014-kz,Kumar2016-xz}, RNA interaction
%% prediction \cite{Pain2015-gr}, protein interactions
%% \cite{Tikk2010-qd}, protein structure prediction
%% \cite{Kolodny2005-ry,Wallner2005-qi}, epistasis \cite{Shang2011-vy},
%% metagenomic analysis \cite{Lindgreen2016-tt,Bazinet2012-wf},
%% repetitive sequence prediction \cite{Saha2008-kd}, proteomics
%% \cite{Lange2008-pt,Yang2009-oc} and phylogenetics
%% \cite{Liu2011-pz,Yang2011-dv,Oscamou2008-md,Bayzid2013-hc,Liu2009-lx}.
Each
software tool was benchmarked in at least one of
\textbf{{\color{red}\numBenchmarkPubs~}} publications that satisfy the Boulesteix
criteria \cite{Boulesteix2013-vb}. In brief, the Boulesteix criteria
are: 1. the main focus of the article is a benchmark. 2. the authors
are reasonably neutral. 3. the test data and evaluation criteria are
sensible.

For each of the publications describing these tools, we have (where
possible) the H5-index published by
\href{https://scholar.google.co.nz/citations?view\_op=top\_venues\&hl=en}{Google
  Scholar Metrics}. We have collected the maximum H-index and
corresponding M-indices \cite{Hirsch2005-mt} for the corresponding
authors for each tool, and the number of times the publication(s)
associated with a tool has been cited using Google Scholar (data
collected over a 6 month period in late 2020). Note that the citation metrics
are not static and will change over time.

We have computed the Spearman’s correlation coefficient for each
pairwise combination of the mean normalised accuracy and speed ranks,
the year published, mean relative age (compared to software in the
same benchmarks), journal H5 metrics, the total number of citations,
the relative number of citations (compared to software in the same
benchmarks) and the maximum H- and corresponding M-indices for the corresponding
authors. Where possible we also extract the version number, the number
of commits and number of contributors from Github repositories.  The
results are presented in Figure~\ref{fig:allfactors}A. We find
significant associations between most of the citation-based metrics
(journal H5, citations, relative citations, H-index and
M-index). There is also a negative correlation between the year of
publication, the relative age and many of the citation-based metrics.

\begin{figure*}[h]
\includegraphics[width=\textwidth]{figure1.pdf}
\caption{\textbf{A.} A heatmap indicating the relationships between
  proposed predictors of software quality. Spearman’s rho is used to
  infer correlations between metrics such as the H- and M-indices of
  corresponding authors, number of citations, journal impact factors
  and H5 indices, the year and relative age of software and the mean
  relative rankings of software speed and accuracy. Red colours
  indicate a positive correlation, blue colours indicate a negative
  correlation. Correlations with a P-value less than 0.05 are
  indicated with a ‘X’. The dashed rectangular area is illustrated in
  more detail in \textbf{B}, the relationship between speed and
  accuracy is shown in more detail in \textbf{Figure 2}.
  %The
  %dendrogram was computed using the default ‘heatmap.2 function in R,
  %which computes a Euclidean distance matrix and a complete-linkage
  %hierarchical clustering.
  \textbf{B.} A barplot illustrating the correlation, as measured by
  Spearman’s rho, between normalised accuracy ranks and software tool
  features that may be predictive of accuracy. In order to give an
  appreciation of the difference between the observed effect-sizes and
  significant effect sizes, we generated 1000 permuted accuracy ranks
  for each benchmark and recorded Spearman’s rho for the significant
  correlations (P<0.05). These values are marked with “x”s in the
  barplot. Significant correlations are marked with red asterisks.}
\label{fig:allfactors}
\end{figure*}

Data on the
number of updates to software tools from Github such as the number of versions,
commits and contributors \textbf{{\color{red}was significantly correlated with 
accuracy (respective Spearman’s rhos = 0.13, 0.22, 0.21, respective P-values = 0.0027, 0.00079, 0.0013). 
These features were not correlated with speed however (see Figure~\ref{fig:allfactors}A).
}}
We also found that author reputation metrics, journal impacts and the age
of tools were generally \textbf{not} correlated with either tool
accuracy or speed (see Figure~\ref{fig:allfactors}). The strongest
association was between accuracy and \textbf{{\color{red}the relative
    number of citations}} (Spearman’s rho = \textbf{{\color{red}0.092,
    P-value = 0.047}}). But the effect size is only just over the
significance threshold and may be explained by multiple-testing.

%To further validate this result, we compute a correlation
%between speed and accuracy for each benchmark and used weighted sum
%Z-tests \cite{Zaykin2011-tj}. This also failed to identify a
%significant relationship (sum Z=-0.1, P-value=0.5).

In order to gain a deeper understanding of the distribution of
available bioinformatic software tools on a speed versus accuracy
landscape, we ran a permutation test. The ranks extracted
from each benchmark were randomly permuted, generating 1,000
randomized speed and accuracy ranks. In the cells of a $10\times10$
grid spanning the normalised speed and accuracy ranks we computed a
Z-score for the observed number of tools in a cell, compared to the
expected distributions generated by 1,000 randomized ranks. The results
of which are shown in Figure~\ref{fig:speedaccuracy}. We identified \textbf{{\color{red}18}}
bins where there was a significant excess or dearth of tools. For
example, there was an excess of “slow and inaccurate” software (\textbf{{\color{red}Z=2.50,
P-value=0.0063}}). We find that the amount of software classed as “fast
and accurate” and “fast and inaccurate” are at approximately the
expected proportions based upon the permutation test. The number of
significant results is in excess, and is not due to multiple testing, as
the probability of finding \textbf{{\color{red}18}} of 100 tests significant by chance is
low (P-value = \textbf{{\color{red}$2.2\times 10^{-6}$}}, exact binomial test).

@@

There is a major reduction in the number of software tools that are
classed as intermediate in terms of both speed and accuracy based on
permutation tests (see Methods for details, Figure~\ref{fig:speedaccuracy}). The cells
corresponding to the four central deciles are
highlighted \textbf{{\color{red}(Z = -1.9, -2.5, -2.9 and -1.1, P-values = 0.03, 0.006,
0.002 and 0.14}}, respectively, reading from top to bottom, left to
right). We also tested the relative age of the software tools in
the indicated corners and central regions of the speed vs
accuracy plot. We found that the “slow and inaccurate” tools ($N=27$) were
generally published earlier than ``fast and accurate'' tools ($N=23$) \textbf{{\color{red}(W = 341, P=0.005,
one-tailed Wilcoxon test)}} (Figure S7).

\section*{Conclusion}

We have gathered data on the relative speeds and accuracies of
\textbf{{\color{red}\numTools~}} bioinformatic tools from
\textbf{{\color{red}\numBenchmarkPubs}} benchmarks published between
2005 and 2020. {\bf Our results suggest there are major benefits for
  long term support for software \cite{siepel2019challenges}.} The
finding of a strong relationship between the number of commits to
github (i.e. software updates) and accuracy, highlights the benefits of
long-term development.

Our study finds little evidence to support that impact-based metrics
are poorly related to software quality, this is unfortunate as these
are frequently cited reasons for selecting software tools
\cite{Loman2015-bw}. This implies that the recorded high citation
rates for bioinformatic software
\cite{Perez-Iratxeta2007-lv,Van_Noorden2014-kc,Wren2016-xy} is more a
reflection of user-friendliness and possibly the Matthew Effect
\cite{Lariviere2010-kx,Merton1968-cb} than accuracy.

We found the lack of a correlation between software speed and accuracy
surprising.  The slower software tools are overrepresented at both
high and low levels of accuracy (Figure~\ref{fig:speedaccuracy}).
In addition, we find an under-representation of software that has
intermediate levels of both accuracy and speed. A possible explanation
for this is that bioinformatic software tools are bound by a form of
publication-bias \cite{Boulesteix2015-am,Nissen:2016}. That is, the
probability that a study being published is influenced by the results
it contains \cite{sterling1959publication}. The community of
developers, reviewers and editors may be unwilling to publish software
that is not highly ranked on speed or accuracy. If correct, this may
have unfortunate consequences.  We also found that slow and inaccurate
software is generally published earlier than fast and accurate tools
\textbf{{\color{red}(P=0.007, one-tailed Wilcoxon test)}}, therefore
comparisons were not required to publish the slow and
inaccurate tools.

%% A
%% gedankenexperiment may be sufficient to show that slow software is
%% less thoroughly tested than fast software. This is because typical
%% academic software development is an iterative process, where tools are
%% refined by successive rounds of coding, testing and evaluation
%% \cite{Wilson2006-ih}. We can assume that similar time-spans are spent
%% on most projects, for example, the span of a PhD degree or
%% Postdoctoral Fellowship.  As a consequence, slow software may undergo
%% fewer development cycles than faster tools.

{\bf STUDY LIMITATIONS...}
